Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-06-14 analysis/paper/paper.Rmd
some updates

Body ornaments are usually viewed as materials that convey social meaning or representations in a culture through the body as the scene to display [@Joyce2005]. Foreign ornaments including glass beads, agate or carnelian beads, and metal objects first appear in Taiwan around 1,800 years ago through regional trade network in Southeast Asia [@WangandLiu2007]. Those new types of ornaments gradually replaced the earlier jade ornamental system in Taiwan during the Metal Age and reached a high peak of quantity and variety in the early 17th century when the Europeans arrived and involved the island in an international trade network on a large scale. Foreign trade goods introduced into the local indigenous societies in an European colonial or imperial contexts usually had an impact on local culture and societies in many parts of the world that usually involved in the transformation of indigenous economic, cultural, and socio-political systems (Dietler 1997; Junker 1993; Mitchell 2000; Silliman 2005). Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed that stress active adoptions of foreign materials, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices (Given 2004; Mullins 2011; Silliman 2001; Rubertone 2000; Torrence 2000; Torrence and Clarke 2000; Voss 2005).

Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare, but the question of long-lasting indirect impacts on local communities remains unanswered. Indirect effects of colonialism is addressed recently to discuss the impact on the indigenous societies in the periphery of colonial control [@Trabert2017]. Taiwan was colonized by the Spanish and the Dutch during the early 17th century, where Northeastern Taiwan is an especially unique example for discussing indirect effects due to its physically isolated location that made it not easily accessible to the Europeans that the colonial control was less compared to other parts of this region (Andrade 2007; Kang 2012). In addition, prior to the European contact, there were small-scale regional exchange networks between China and Taiwan. Also, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and encountered a large wave of Han Chinese migration in the 19th century. This made northeastern Taiwan a good case to explore culture contacts at different periods and their indirect impacts by taking a long-term perspective.

We ask whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan.  Our hypothesis is that there was a greater diversity of ornaments types and materials at Kiwulan after European contact. We hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan. We hypothesize that there is no change in the spatial distribution of ornaments at Kiwulan relating to these international contacts.

@@

@@

The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified from Chinese official records [b]and large amount of Chinese blue-and-white porcelain found at Kiwulan.[c]

The ornaments sampled in this study come from 40 adjacent 4m x 4m units out of 262 units in total since those sampled units were less disturbed by modern constructions and remains intact. The Upper Layer Culture can be divided into six sub layers spanning from 14 century to 19 century according to current radiocarbon dates, excavation depth, types of ceramics, and sediment texture and color (Hsieh 2008; Wang 2011). Layers are assigned numbers from 1 to 6 for which L1 corresponds to the upper context and L6, also the bottom layer, refers to lower context. Based on Hsieh’s (2008) chronological study of Chinese porcelains, L5 and L6 represent pre-European contact period, L4 was the time of European contact, and L2 and L1 was the period of Chinese contact.

Graph of 40 units and chronology

@@

To explore whether body ornaments can reflect the colonial influences on indigenous populations, this research examines the frequency and spatial distribution of trade ornaments at Kiwulan during different time periods with foreign contacts, including European and Chinese contacts. Trade ornaments were analyzed by three analytical units, including pre-European contact, European contact, and Chinese contact. Because the appearance of ornament is highly correlated to the raw material itself, which is selected as the first criteria for classification. Under each category of raw material, ornaments are further classified into subtypes according to shapes when variation are found in each category. To compare the general pattern of changes in major ornaments, less frequent ornaments are combined as one category named other, which will be discussed separately.

There are `r sample_size` ornaments excavated from the 40 sampling squares, which accounts for `r sample_pro`% of the total number of ornaments found in the excavation section where they were from. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. Ornaments in some categories are divided into sub-types due to the variations of shape, including agate bead, metal ring, and metal bell. The frequency of the subtypes are summarized in \@ref(tab:tidy-data-for-table).

The results of the comparison between different time periods show a general pattern that most ornaments appeared in the pre-European contact period and the frequencies reached a peak during the European contact and then dropped during the Chinese contact period, especially for golden bead. This trend can be also seen on other ornaments including agate bead, metal ring, and bell. However, glass bead shows a different pattern that indicates a higher frequency in the pre-European contact, and then the frequency decreased in the European contact period and even less in the Chinese contact period.       
The distribution of frequency for subtypes in each major category are presented in (make a figure). Agate bead can be divided into 7 subtypes, in which hexagonal shape is the most common subtype that appeared before European contact and the number increased in the Eruopean contact and then declined in the Chinese contact. Small oval type shows similar frequency before and during European contact but not found in the Chinese contact period, while waxy oval bead only found during European contact period. There are more subtypes during European contact during the European contact compared to other time periods.

For golden bead, we focus on their length that

2. Ornaments by: 
 - time period

@@

@@

```{r scatter-plot-bead-by-period}
# plot for intact golden bead by period
ornaments_intact_gold <-
  ornaments_period %>% 
  filter(Condition %in% c('complete', 'near-complete'),
         Categories == 'Golden bead')

@@

```{r plot-count-type-by-period}
# plot for different shape of agate bead by period
agate_bead_shape <-
  ornaments_period %>% 
  filter(Categories == 'Agate bead') %>%
  ungroup() %>% 
  mutate(Type = ifelse(Type == 'prism/octagonal', 
                       str_remove(Type, '/octagonal'), Type)) %>% 
  mutate(Type = ifelse(Type == 'prism/pentagonal', 
                       str_remove(Type, '/pentagonal'), Type)) %>% 
  mutate(Type = ifelse(Type == 'prism/hexagonal', 
                       str_remove(Type, 'prism/'), Type)) %>%
  mutate(Type = fct_infreq(Type)) %>%
  mutate(Type = fct_rev(Type)) %>% 
  group_by(period, Type) %>% 
  count()

>>>>

Body ornaments are usually viewed as materials that convey social meaning or representations in a culture through the body as the scene to display [@Joyce2005]. Foreign ornaments including glass beads, agate or carnelian beads, and metal objects first appeared in Taiwan around 1,800 years ago through the regional exchange network in Southeast Asia [@WangandLiu2007]. Those new types of ornaments gradually replaced the earlier jade ornamental system in Taiwan since the transition of Late Neolithic and the Iron Age. We observed a greater quantity and variety of foreign goods in the early 17th century during the European canonization period that involved the whole island in an international trade network on a large scale. Foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts usually had an impact on local societies in many parts of the world that might cause transformations of indigenous economic, cultural, and socio-political systems (Dietler 1997; Junker 1993; Mitchell 2000; Silliman 2005). Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed that emphasize the ability of active adoptions of foreign materials, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices (Given 2004; Mullins 2011; Silliman 2001; Rubertone 2000; Torrence 2000; Torrence and Clarke 2000; Voss 2005).

Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local communities remains unanswered. Indirect effects of colonialism is addressed recently to discuss the impact on the local indigenous societies in the periphery of colonial control [@Trabert2017]. Taiwan was colonized by the Spanish and the Dutch during the early 17th century, where northeastern Taiwan, Yilan, is an especially unique example for exploring indirect effects due to its physically isolated location that made it not easily accessible to the Europeans and the colonial control was less compared to other parts of this region (Andrade 2007; Kang 2012). Prior to the contact with the Europeans, there were small-scale regional exchange networks between China and Taiwan since in the Neolithic period. Also, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and encountered a large wave of Han Chinese migration in the 19th century. Northeastern Taiwan can provide a good example to explore the indirect impacts on indigenous societies caused by different foreign groups in different periods by taking a long-term perspective research.

We ask whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan. Our hypothesis is that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network. We hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan. We hypothesize that there is no change in the spatial distribution of ornaments at Kiwulan relating to these international contacts.

Q: Is the difference in the frequency of ornaments related to the scale of exchange network, does it show any spatial pattern at Kiwulan that indicate the appearance of social inequality.

Q: what is the scale of those exchange network: 
- pre-European contact: middle scale, regional trade network 
- European contact: global scale, international trade network 
- Chinsese contact: small scale, mutual trade network

Q: Is the difference in the frequency of ornaments related to the intensity of contact, does it show any spatial pattern at Kiwulan that indicate the appearance of social inequality.
- pre-European contact: indirect 
- European contact: indirect, colonial context
- Chinsese contact: direct, Han people migration

 and intense of contact

@@

<--Add a paragraph about exchange network before European contact-->

@@

The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified from Chinese official records [b]and large amount of Chinese blue-and-white porcelain found at Kiwulan.

The ornaments studied in this paper came from 40 adjacent 4m x 4m units out of 262 units in the middle part of excavation. The sampling principle is based on the extent of intact contexts that the least disturbed by modern constructions are the preference. The Upper Layer Culture can be divided into six sub layers spanning from 14 century to 19 century according to current radiocarbon dates, excavation depth, types of ceramics, and sediment texture and color (Hsieh 2008; Wang 2011). Layers are assigned numbers from 1 to 6 for which L1 corresponds to the upper context and L6, also the bottom layer, refers to lower context. Based on Hsieh’s (2008) chronological study of Chinese porcelains, L5 and L6 represent pre-European contact period, L4 was the time of European contact, and L2 and L1 was the period of Chinese contact.

<--add a graph of 40 units and chronology-->

@@

@@

To explore whether body ornaments can reflect the colonial influences on indigenous populations, this research examines the frequency and spatial distribution of trade ornaments at Kiwulan during different time periods with foreign contacts, including European and Chinese contacts. Trade ornaments were analyzed by three analytical units including pre-European contact, European contact, and Chinese contact that indicates three scale of exchange network from less to intense contact subsequently. Because the appearance of ornaments is highly related to the original raw material couple with the function such as bead and bracelet, they are combined as the first criterion of classification which determines five major categories. Under each category of raw material, ornaments are further classified into several subtype according to the shape if any variations were identified in each category. In order to compare the general pattern of changes for major ornaments, less frequent ornaments are combined as one category named other, which will be discussed separately.

Ornaments with clear contexts were mainly found from the post-hole area, burials, and middens, where ornaments from the post-hole area accounts for the majority of the amount in total. Post-hole area is also the contexts that we mainly focus on because it represents house units accross the site that can further provide a basis for comparison to see if there is any spatial pattern of distribution. Ornaments from middens and burials will also be discussed in the later section. For the post-hole area, there were `r sample_size` ornaments recovered from the 40 sampling squares, which accounts for `r sample_pro`% of the total number of ornaments found in the same excavation section including all contexts. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. Ornaments in most categories can be divided into sub-types according to the variation of shape. The variation was defined by the difference identified in at least two variables, for example dimensional variables such as length, or stylish variables such as patterns. The frequency and feature of subtypes for major categories are summarized in \@ref(tab:tidy-data-for-table).

The results of the comparison between different time periods show a general pattern that most ornaments appeared in the pre-European contact period and the frequencies reached a peak during the European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate bead, metal ring, and bell. However, glass bead shows a different pattern that indicates a higher frequency in the pre-European contact, and then the frequency decreased in the European contact period and even less in the Chinese contact period.

The distribution of frequency for subtypes in each major category are presented in (make a figure to combine all). Agate beads can be divided into 7 subtypes, in which hexagonal shape is the most common subtype that appeared before European contact and the number increased in the Eruopean contact and then declined in the Chinese contact. Small oval type shows similar frequency before and during European contact but not found in the Chinese contact period, while waxy oval bead only found during European contact period. There are more subtypes of agate beads during European contact compared to other time periods. For golden bead, we focus on their length

## Frequency for spatial pattern
2. Ornaments by: 
 - time period

@@

```{r scatter-plot-bead-by-perio, eval= FALSE}
# plot for intact golden bead by period
ornaments_intact_gold <-
  ornaments_period %>% 
  filter(Condition %in% c('complete', 'near-complete'),
         Categories == 'Golden bead')

@@

```{r plot-count-type-by-period}
# tidy the data with categories and shapes
ornaments_shape_count <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring",
                            "Golden bead"))

# plot for different shape of agate bead by period
agate_bead_shape <-
  ornaments_period %>% 
  filter(Categories == 'Agate bead') %>%
  ungroup() %>% 
  mutate(Type = ifelse(Type == 'prism/octagonal', 
                       str_remove(Type, '/octagonal'), Type)) %>% 
  mutate(Type = ifelse(Type == 'prism/pentagonal', 
                       str_remove(Type, '/pentagonal'), Type)) %>% 
  mutate(Type = ifelse(Type == 'prism/hexagonal', 
                       str_remove(Type, 'prism/'), Type)) %>%
  mutate(Type = fct_infreq(Type)) %>%
  mutate(Type = fct_rev(Type)) %>% 
  group_by(period, Type) %>% 
  count()
